ODGI: understanding pangenome graphs

### 2021-10-30 manuscript/main.tex
current subsection 4.2 is tricky

\begin{itemize}
\item \textit{odgi build} constructs the ODGI data model from GFA (\S\ref{sec:build}).
\item \textit{odgi view} converts the ODGI data model into GFA (\S\ref{sec:text}).
\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item \textit{odgi position} lifts coordinates between paths and graph positions (\S\ref{sec:untangle}).
\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

Each tool focuses on a small set of related operations.
Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
This supports the implementation of flexible and composable graph processing pipelines based on a common data type.
While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
Despite being recent systems, pangenome graphs are not incompatible with current standard ways of working with bioinformatics data.
ODGI provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
Indeed, by considering all paths in the graph as potential reference or query sequence, we make graphs invisible to downstream tools that operate on collections of sequences or rely on a reference sequence (\textit{e.g.} SAMtools~\citep{Li2009}), enabling interoperability.
This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

@@

\subsection{Building the \textsc{ODGI} model}
\label{sec:build}

We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA. Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes. Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure \ref{fig:build}).

%Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)

\subsection{Converting to standard file formats}
\label{sec:text}

% pjotr & andrea
ODGI maintains its own efficient binary format for storing graphs on disk.
ODGI also supports reading and writing the textual GFA file format for interchange with other tools such as with the VG toolkit~\citep{Garrison:2018}.

\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1~\citep{GFA} format. It can reveal a graph’s internal structures for e.g. debugging processes.

\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph. {\color{red} FIXME: All this information is also listed in the metrics section. How to split? No mentioning of file formats in the following sections?}

@@

@@

\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.

@@

@@

% erik

@@

% simon

@@

% erik & pjotr & anyone

>>>>

\begin{itemize}
\item \textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
\item \textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
\item \textit{odgi position} lifts coordinates between paths and graph positions (\S\ref{sec:untangle}).
\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
\end{itemize}

Each tool focuses on a small set of related operations.
Most read or write the native ODGI format (`og' extension) (Figure~\ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
This supports the implementation of flexible and composable graph processing pipelines based on a common data type.
While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
Despite being recent systems, pangenome graphs are not incompatible with current standard ways of working with bioinformatics data.
ODGI provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
Indeed, by considering all paths in the graph as potential reference or query sequence, we make graphs invisible to downstream tools that operate on collections of sequences or rely on a reference sequence (\textit{e.g.} SAMtools~\citep{Li2009}), enabling interoperability.
This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

@@

\subsection{Building the \textsc{ODGI} model}
\label{sec:build}
ODGI maintains its own efficient ODGI binary format for storing graphs on disk.
We begin by transforming the storage model of the standard GFAv1~\citep{GFA} format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
This construction step can be a significant bottleneck, in particular as the size of the path set of the graph increases.

The ODGI data structure (Algorithm~\ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA file.
Then, we serially add edges in the $Node.edges$ vectors of pairs of nodes.
Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure~\ref{fig:build}).

For interchange with other tools, \textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1~\citep{GFA}.

\subsection{Converting to standard file formats{\color{red} FIXME: We need a different title/subsection}
}
\label{sec:text}

@@

{\color{red} FIXME: We should put the following information in the metric section, of move some commands here}

@@

\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently. {\color{red} FIXME: this command is listed in the list in 4 Results}

@@

\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.

@@

@@

@@
